---
title: "Distribution of Random Walks"
output: rmarkdown::tufte_handout
---

## Convergence of Binomial Probabilities^[These examples are based on: http://nvenkataraman1.github.io/random-walks/]

We all intuitively understand that when a fair coin (50% heads, 50% tails) is flipped many times. the probability of heads versus tails should start to converge. For example, let's say we get 1 point for every head and 0 points for every tail.  What will the *average* number of points we get at any point in time?  Let's compute and plot this in R with `n=5000` coin flips!
```{r echo=FALSE}
set.seed(50)
```
```{r plot_binom_convergence, fig.cap = "Coin flipping probabilities", dpi=150, fig.margin=TRUE, fig.width=5, fig.height=4, cache=TRUE}
n <- 5000
flips <- rbinom(n, 1, 0.5)    # a fair coin follows a simple binomial distribution 
means <- c(0)
for (i in 1:n) {
  means[i] <- mean(flips[1:i])
}

plot(means, type="l", ylim=c(0,1))
abline(h=0.5, col="red")
```

We see that the average number of points quickly converges to 0.5 (a *half*), just as we expected. This is our common intuition about how probabilities average out over time.

\newthought{Now, instead of average probabilities}, let's play a game where we gain one point for heads and lose one point for tails.  How many *total points* would we make over time? Will it converge?
```{r echo=FALSE}
set.seed(60)
```
```{r plot_random_binom_walk, fig.cap = "A random (binomial) walk", dpi=150, fig.margin=TRUE, fig.height=4, cache=TRUE}
n <- 5000
flip <- 2*rbinom(n, 1, 0.5)-1
total <- c(0)
for (i in 2:n) { total[i] <- total[i-1] + flip[i] }

plot(total, type="l", ylim=c(-300, 300))
abline(h=0)
```

When we plot out our total number of points, rather than converging, it seems to just randomly walk away! *Random walks* are one way of modeling many everyday phenomena that are affected by so many variables that they appear random in their progress: stock markets, weather fluctuations, ant foraging, brownian motion, and more^[http://en.wikipedia.org/wiki/Random_walk].

\newthought{But how far from their starting points} do random walks go *on average*? It isn't easy to tell, so let's simulate random walks and find out!
\marginnote{We first create a `function` that creates binomial walks for us. It reuses the code we saw earlier. This function takes the number of `steps` to walk as a parameter.}
```{r echo=FALSE}
set.seed(100)
```
```{r binomial_walk_function, cache=TRUE} 
# Random binomial walk function
random_binomial_walk <- function(steps) {
  flip <- 2*rbinom(n, 1, 0.5)-1
  total <- c(0)
  for (i in 2:steps) { total[i] <- total[i-1] + flip[i] }
  return(total)
}
```
\marginnote{We first create a `data.frame` that is the size of a 2x3 matrix. The we use a loop to run our random binomial walk function six times. The result of our six function calls are stored in the data.frame

We then plot the six walks. The `mfrow` function lets us put plots side-by-side in a grid pattern! At the end, we have to be careful to call mfrow again to return the graphics setting to plot 1x1 figures at a time.}
```{r plot_six_binomial_walks, fig.cap = "Six Random Binomial Walks", fig.width=7, fig.height=4.5, cache=TRUE}
# Simulate six random walks of 5000 steps each
walks <- data.frame(matrix(0, nrow = 5000, ncol = 6))
for (i in 1:6) { 
  walks[i] <- random_binomial_walk(5000);
}

# Plot out our six random walks
par(mfrow=c(2,3)) # setup plotting to 2rows x 3 columns
for (i in 1:6) { 
plot(walks[[i]], type="l", ylim=c(-200, 200))
abline(h=0, col="red")
}
par(mfrow=c(1,1)) # return graphics settings to 1 x 1
```

Even with a small number of random walks, we see they can go anywhere. It will be hard to plot a large number of walks. So let's simulate a large number of walks and see what distance from zero points each walk ended up. We will use m=3000 walks, each having n=2000 steps.
```{r dist_binomial_walks, cache=TRUE}
m <- 3000
n <- 2000
walks <- data.frame(matrix(0, nrow = n, ncol = m))
for (i in 1:m) { walks[i] = random_binomial_walk(n) }
final.points <- c(0)
for (i in 1:m) { final.points[i] <- walks[n,i] }
```

We can now plot the distribution of how many points away from zero (positive or negative) each each of the 3000 walks finally got to, after 2000 steps.
```{r plot_dist_binomial_walks, fig.cap = "Distribution of binomial walks", fig.height=4, fig.margin=TRUE, cache=TRUE}
hist(final.points, breaks=20, prob=TRUE)
lines(density(final.points), lwd=3, col="blue")
```

Does this bell-shaped symmetric distribution look familiar? It is our old friend the *normal distribution* again!