---
title: "Sampling and Inference"
author: "Soumya Ray"
date: "Oct 11, 2014"
output: rmarkdown::tufte_handout
---

# Sampling Process and Inferential Error
We can create a population and then create a noisy sampling frame. This increases the 'dispersion' of the sample by adding noise to it.

\marginnote{This code block creates a population and noise vectors of size N, creates a sampling frame with noise, and samples from it.}
```{r}
N=10000
n=150
pop <- rnorm(N, mean=22, sd=5)
noise <- rnorm( length(pop), sd=4 )
noisy.frame <- pop + noise
noisy.sample <- sample(noisy.frame, size=n)
```

We can now use the `plot` command to visualize the dispersion of the original population (solid curve), compared to that of the sample (dashed):
```{r plot, fig.cap = "plot figure", dpi=150, fig.margin=TRUE, fig.width=5, fig.height=4}
# Plot the population and noisy sampling frame
plot(density(pop),  main=NULL)
lines(density(noisy.sample), lty="dashed")
```
  
  
  
And now the same plot using `ggplot`. This time we have to create data.frame objects of the population and sample data:
```{r, echo=FALSE}
library(ggplot2)
```

```{r ggplot, fig.cap = "ggplot figure", fig.margin=TRUE}
pop.df = as.data.frame(pop)
sample.df = as.data.frame(noisy.sample)
ggplot() + geom_density(data=pop.df, aes(pop)) + 
           geom_density(data=sample.df, 
                        aes(noisy.sample), linetype="dashed")
```










